\documentclass[11pt]{article}

\usepackage[margin=1in]{geometry}
\usepackage{amsmath}
\usepackage{booktabs}
\usepackage{graphicx}
\usepackage{hyperref}
\usepackage{siunitx}
\usepackage{pgfplots}
\usepackage{tikz}
\usepgfplotslibrary{groupplots}
\pgfplotsset{compat=newest}

\title{Reproducing LiteFNO on Gray-Scott Reaction--Diffusion: A Detailed Study for Low-Resource Scientific Modeling}
\author{LiteFNO Reproduction Team}
\date{\today}

\begin{document}
\maketitle

\begin{abstract}
This paper presents a detailed reproduction of the Lightweight Fourier Neural
Operator (LiteFNO) for time-dependent partial differential equations, following
the NeurIPS ML4PS 2025 paper ``Lightweight Fourier Neural Operator for
Time-Dependent Partial Differential Equations'' \cite{litefno2025}. We evaluate
LiteFNO on the Gray-Scott reaction--diffusion dataset from The Well benchmark,
using the configuration defined for this study. The model attains a best
validation RMSE of \num{0.013112} and validation VRMSE of \num{0.034148} at epoch
29, with a test RMSE of \num{0.013201} and test VRMSE of \num{0.035189}. The run
spans 33 logged epochs (0--32) and uses a 107,842-parameter model. Beyond a
metric summary, we document preprocessing, training details, evaluation
protocols, and learning curves, and we discuss limitations and extension
opportunities in low-resource settings.
\end{abstract}

\section{Introduction and Motivation}
Neural operators provide a data-driven approach to learning solution operators
for time-dependent partial differential equations (PDEs), enabling rapid
surrogate inference once trained. Fourier Neural Operators (FNOs) introduced a
powerful spectral formulation for learning operators in function spaces
\cite{li2020fourier}. LiteFNO extends this direction with a factorized
spatio-temporal formulation that substantially reduces parameters and improves
efficiency \cite{litefno2025}. This paper reproduces the LiteFNO results on the
Gray-Scott reaction--diffusion task, providing a transparent and detailed
account of data handling, model configuration, and observed metrics.

The motivation extends beyond benchmarking. Climate adaptation decisions for
smallholder farmers in Africa require localized forecasting and scenario
analysis under tight computational and infrastructure constraints. Low-footprint
surrogate models can support rapid exploration of ``what-if'' scenarios for
water availability, planting schedules, and climate stressors without requiring
large-scale HPC resources. By reproducing LiteFNO in a controlled setting, we
assess its potential as a building block for deployable scientific ML pipelines
that can operate at the edge or in bandwidth-limited environments.

\section{Related Work}
Fourier Neural Operators (FNOs) \cite{li2020fourier} represent a seminal class of
operator-learning models that apply spectral convolutions to learn mappings
between function spaces. Extensions of FNOs for temporal modeling typically
handle time either autoregressively in the spatial domain or by lifting the
entire space--time grid into spectral space, trading off computational cost and
modeling fidelity. LiteFNO \cite{litefno2025} introduces a low-rank factorization
strategy for spatial spectral weights and a transductive temporal basis to
recover spatio-temporal interactions with far fewer parameters. This
reproduction follows the LiteFNO configuration and evaluation protocols to
provide an independent account of performance on the Gray-Scott benchmark.

\section{Background}
Neural operators aim to learn mappings \(\mathcal{G}: \mathcal{A} \to \mathcal{U}\)
between function spaces, such as from initial conditions to PDE solutions.
In Fourier Neural Operators, a typical layer updates a latent field \(v\) via
spectral convolution,
\begin{align}
v_{k+1}(x) = W v_k(x) + \mathcal{F}^{-1} \left( R \cdot \mathcal{F}(v_k)(\xi) \right),
\end{align}
where \(\mathcal{F}\) denotes the Fourier transform, \(R\) is a learned complex
filter (often truncated to low modes), and \(W\) is a local linear transform.
LiteFNO factorizes the spectral weights into a low-rank spatial basis and a
temporal basis learned via transduction, enabling efficient modeling of
spatio-temporal dynamics with a fraction of the parameters \cite{litefno2025}.

\section{Method details: LiteFNO}
LiteFNO addresses the trade-off between spatio-temporal expressivity and
computational cost by factorizing the spectral weights into structured bases.
Concretely, the spatial spectral kernel is parameterized using a low-rank basis,
which reduces the number of learned coefficients while still capturing global
interactions. Temporal dynamics are then incorporated by learning a temporal
basis and coupling it with the spatial basis through transduction. This
formulation approximates a full space--time operator without explicitly learning
a dense spatio-temporal kernel, offering a practical compromise between the
autoregressive spatial approach and the fully spatio-temporal spectral approach.
In this reproduction, the architecture is used as described in
\cite{litefno2025} without any architectural modifications.

\section{Dataset and preprocessing}
The evaluation uses the Gray-Scott reaction--diffusion dataset from The Well
benchmark, following the dataset usage described in the LiteFNO paper
\cite{litefno2025}. The raw data are HDF5 arrays with shape
\((n_{\text{traj}}, n_{\text{steps}}, H, W, \text{fields})\). Preprocessing
downsamples spatially, caps the number of trajectories, and limits the temporal
extent to align with the training regime. Table~\ref{tab:dataset} summarizes the
exact settings used in this reproduction.

The Gray-Scott system is a canonical reaction--diffusion model that generates
rich spatio-temporal patterns. It serves as a controlled benchmark for
evaluating whether an operator model can capture non-linear diffusion-driven
structure while remaining stable over time. The dataset is split into train,
validation, and test partitions, and the preprocessing pipeline applies a
consistent trajectory cap to reduce the training footprint while preserving
diverse regimes.

\begin{table}[t]
  \centering
  \caption{Gray-Scott preprocessing configuration.}
  \label{tab:dataset}
  \begin{tabular}{ll}
    \toprule
    Setting & Value \\
    \midrule
    Dataset name & \texttt{gray\_scott\_reaction\_diffusion} \\
    Downsample factor & 4 \\
    Max trajectories & 1000 \\
    Max steps & 60 \\
    Input steps & 1 \\
    Output steps & 1 \\
    Fields & 2 \\
    Stride & 1 \\
    Cache & \texttt{memory} \\
    Random seed & 0 \\
    Dataset key & \texttt{data} \\
    \bottomrule
  \end{tabular}
\end{table}

Planned extensions include low-rank and width sweeps, quantization-aware
training, robustness checks under distribution shift, and explainability
analyses to attribute spectral modes to spatial patterns. These extensions align
with the low-resource deployment objective but are not yet executed in the
current log, so no extension-specific quantitative results are reported here.
We also plan to investigate the effect of input--output horizon length on
multi-step forecasting performance, which will enable windowed VRMSE reporting
consistent with \cite{litefno2025}. Another extension is to evaluate sensitivity
to the number of retained Fourier modes, which can further trade accuracy for
inference speed in deployment settings.

\section{Model and training setup}
LiteFNO uses a compact spectral architecture configured via a set of experiment
parameters. The Gray-Scott run adopts the base LiteFNO model with a dataset
override for preprocessing and logging. The training configuration includes
AdamW-style optimization with a step learning-rate schedule, mixed-precision
training (AMP), and evaluation on both train and validation splits. Core
settings are listed in Table~\ref{tab:training}.

\section{Implementation and logging}
Training and evaluation are instrumented through a metrics logger that records
per-epoch RMSE and VRMSE for each evaluated split. Logs are serialized as JSONL,
one record per epoch, enabling straightforward ingestion for plotting and
analysis. The experiment specifies evaluation on both train and validation
splits at every epoch and triggers an explicit test evaluation after training.
Checkpoints are configured to track the best model according to validation
VRMSE, which is the primary selection criterion in this study.

To minimize memory pressure and improve throughput on limited hardware, mixed
precision (AMP) is enabled. The configuration also uses pinned memory and sets
the number of data-loader workers to zero, which simplifies reproducibility on
machines without extensive multiprocessing support.

\begin{table}[t]
  \centering
  \caption{Model and training configuration for the Gray-Scott LiteFNO run.}
  \label{tab:training}
  \begin{tabular}{ll}
    \toprule
    Setting & Value \\
    \midrule
    Layers & 8 \\
    Width & 64 \\
    Rank & 32 \\
    Epochs (configured) & 200 \\
    Batch size & 512 \\
    Learning rate & 1e{-}3 \\
    LR step & 100 \\
    LR gamma & 0.5 \\
    Eval every & 1 \\
    Eval splits & train, valid \\
    Eval windows & [6, 12], [13, 30] \\
    Test at end & true \\
    Device & \texttt{cuda} \\
    Seed & 1337 \\
    AMP & true \\
    Checkpoint best metric & \texttt{valid\_vrmse} \\
    \bottomrule
  \end{tabular}
\end{table}

\section{Metrics}
We report root mean squared error (RMSE) and variance-normalized RMSE (VRMSE).
For prediction \(\hat{y}\) and target \(y\), the metrics are:
\begin{align}
\mathrm{RMSE} &= \sqrt{\mathbb{E}\left[(\hat{y} - y)^2\right]}, \\
\mathrm{VRMSE} &= \sqrt{\frac{\mathbb{E}\left[(\hat{y} - y)^2\right]}{\mathrm{Var}(y) + \epsilon}},
\end{align}
where \(\epsilon\) is a small stabilizer.

In addition to scalar RMSE and VRMSE, the configuration specifies evaluation
windows \([6, 12]\) and \([13, 30]\) to support windowed VRMSE analysis when the
prediction horizon is sufficiently long. The current log does not include
windowed metrics because the run uses single-step output; nevertheless, these
definitions are retained for compatibility with longer-horizon extensions.

\section{Reproduction details}
The reproduction run follows the end-to-end workflow outlined in
\cite{litefno2025} and is executed using a consistent configuration for
preprocessing, training, and evaluation. The core workflow is:
\begin{verbatim}
litefno download --config configs/datasets/gray_scott_reaction_diffusion.yaml
litefno preprocess --config configs/datasets/gray_scott_reaction_diffusion.yaml
litefno train --config configs/experiments/litefno_gray_scott_reaction_diffusion.yaml
litefno test --config configs/experiments/litefno_gray_scott_reaction_diffusion.yaml \
  --checkpoint outputs/checkpoints/gray_scott_reaction_diffusion/litefno/best.pt
\end{verbatim}
Metrics are logged as JSONL records to
\texttt{outputs/logs/gray\_scott\_reaction\_diffusion\_litefno.jsonl}. The
learning curves in Figure~\ref{fig:learning-curves} are generated from the
recorded log in this study at
\texttt{logs/gray\_scott\_reaction\_diffusion\_litefno.jsonl}.

The reproduction tracks both the training trajectory and the validation
trajectory, allowing inspection of overfitting and generalization behavior. We
use the best validation VRMSE checkpoint for final test evaluation, mirroring
the model-selection procedure described in \cite{litefno2025}. This avoids
reporting a test metric from an arbitrary epoch and provides a consistent
selection rule for future extensions.

\section{Baselines and comparisons}
This study reports a single LiteFNO reproduction run and does not include a
direct baseline. In the original LiteFNO paper, comparisons are made against
FNO variants that either model time autoregressively or operate on the full
spatio-temporal grid \cite{litefno2025}. A planned next step is to run a
matching FNO-S baseline on the same Gray-Scott preprocessing pipeline to
quantify the parameter--accuracy trade-offs under identical conditions.

\section{Extensions \& Results}
This section reports the baseline reproduction results from the existing log
and outlines planned extensions for low-resource deployment.

The training log (\texttt{logs/gray\_scott\_reaction\_diffusion\_litefno.jsonl})
contains 33 epochs (0--32) plus a test evaluation recorded with step \(-1\).
Validation metrics improve rapidly early in training and plateau by the late
epochs. Table~\ref{tab:results} lists the best validation and test performance
and the logged parameter count.

\begin{table}[t]
  \centering
  \caption{Summary of Gray-Scott LiteFNO results.}
  \label{tab:results}
  \begin{tabular}{ll}
    \toprule
    Metric & Value \\
    \midrule
    Best validation epoch & 29 \\
    Best validation RMSE & \num{0.013112} \\
    Best validation VRMSE & \num{0.034148} \\
    Final logged epoch & 32 \\
    Final validation RMSE & \num{0.013310} \\
    Final validation VRMSE & \num{0.034442} \\
    Test RMSE & \num{0.013201} \\
    Test VRMSE & \num{0.035189} \\
    Parameters & 107,842 \\
    \bottomrule
  \end{tabular}
\end{table}

\section{Learning dynamics and error characteristics}
The learning curves in Figure~\ref{fig:learning-curves} show steep improvement
over the first several epochs, followed by a gradual flattening. There is mild
non-monotonicity in both RMSE and VRMSE, which is typical of stochastic
optimization with periodic learning-rate decay. The best validation metrics
occur slightly before the final logged epoch, suggesting that early stopping or
more frequent checkpoint selection could marginally improve final test scores.

From a numerical standpoint, VRMSE tracks RMSE closely because the variance of
the target fields is stable across epochs. This makes VRMSE a convenient
normalized metric for comparing across datasets with different scales. The
small gap between validation and test VRMSE indicates that the model's
generalization error remains controlled under the current preprocessing regime.

\begin{figure}[t]
  \centering
  \begin{tikzpicture}
    \begin{groupplot}[
      group style={group size=2 by 1, horizontal sep=1.2cm},
      width=0.48\textwidth,
      height=0.35\textwidth,
      xlabel={Epoch},
      grid=both,
      grid style={line width=.1pt, draw=gray!30}
    ]
      \nextgroupplot[ylabel={RMSE}]
      \addplot[blue] table [x=step, y=train_rmse, col sep=comma] {data/gray_scott_litefno_metrics.csv};
      \addplot[orange] table [x=step, y=valid_rmse, col sep=comma] {data/gray_scott_litefno_metrics.csv};
      \legend{Train,Valid}
      \nextgroupplot[ylabel={VRMSE}]
      \addplot[blue] table [x=step, y=train_vrmse, col sep=comma] {data/gray_scott_litefno_metrics.csv};
      \addplot[orange] table [x=step, y=valid_vrmse, col sep=comma] {data/gray_scott_litefno_metrics.csv};
      \legend{Train,Valid}
    \end{groupplot}
  \end{tikzpicture}
  \caption{Training and validation metrics for LiteFNO on Gray-Scott.}
  \label{fig:learning-curves}
\end{figure}

The gap between validation and test performance is small, suggesting good
generalization for this dataset under the configured preprocessing. The best
validation epoch (29) slightly outperforms the final epoch (32), indicating
modest metric oscillations late in training. Despite a compact parameter count,
LiteFNO achieves low VRMSE relative to the initial epoch, with validation RMSE
improving by roughly 70\% from epoch 0 to the best epoch.

Planned extensions include low-rank and width sweeps, quantization-aware
training, robustness checks under distribution shift, and explainability
analyses to attribute spectral modes to spatial patterns. These extensions align
with the low-resource deployment objective but are not yet executed in the
current log, so no extension-specific quantitative results are reported here.

\section{Limitations}
This report is limited to a single LiteFNO run on one dataset and does not yet
include a baseline comparison (e.g., FNO-S) or multiple random seeds. The log
contains only 33 epochs, which is shorter than the 200 epochs configured in the
experiment, so the results should be interpreted as an early or partial run.
Windowed VRMSE metrics are not available in the current log, and the planned
extensions have not yet produced results.

\section{Reproducibility checklist}
Table~\ref{tab:repro} summarizes the minimal information required to reproduce
the current results. This checklist is intended to support transparent reporting
and straightforward re-execution of the experiment.

\begin{table}[t]
  \centering
  \caption{Reproducibility checklist for the LiteFNO Gray-Scott run.}
  \label{tab:repro}
  \begin{tabular}{ll}
    \toprule
    Item & Value \\
    \midrule
    Dataset & Gray-Scott reaction--diffusion (The Well) \\
    Data splits & train / valid / test \\
    Preprocessing & downsample=4, max\_trajectories=1000, max\_steps=60 \\
    Model & LiteFNO (layers=8, width=64, rank=32) \\
    Training & epochs=200 (configured), batch\_size=512, lr=1e{-}3 \\
    Scheduler & step=100, gamma=0.5 \\
    Seed & 1337 \\
    Metrics & RMSE, VRMSE \\
    Best model selection & validation VRMSE \\
    Metrics log & outputs/logs/gray\_scott\_reaction\_diffusion\_litefno.jsonl \\
    \bottomrule
  \end{tabular}
\end{table}

\section{Future work}
Future work will complete baseline comparisons and evaluate robustness across
multiple seeds. We will extend the analysis to longer prediction horizons to
enable windowed VRMSE reporting and will explore model compression techniques
that preserve accuracy under quantization. Finally, we plan to validate the
utility of LiteFNO-style surrogates on downstream climate-adaptation tasks,
including localized water-stress forecasting and scenario planning for
smallholder agriculture.

\section{Deployment considerations for climate adaptation}
Low-resource deployment contexts place practical constraints on model size,
training cost, and inference latency. LiteFNO is designed to reduce parameters
through low-rank spectral factorization while maintaining accuracy, making it a
reasonable candidate for embedded or cloud-edge deployments. In agricultural
settings, the ability to run rapid surrogate simulations can enable interactive
decision support, such as stress-testing planting schedules or exploring
irrigation strategies under projected climate scenarios. While the current
experiment is limited to a canonical PDE benchmark, it provides quantitative
evidence that compact neural operators can achieve competitive accuracy with
small memory footprints, which is a necessary precondition for deployment in
resource-constrained environments.

\section{Ethical and societal considerations}
Decision-support systems in climate adaptation must be deployed cautiously,
particularly in contexts where data availability, infrastructure, and local
agronomic knowledge vary widely. Lightweight models can lower barriers to
adoption, but they also risk being applied outside the data regimes they were
trained on. Any downstream application should pair model predictions with
uncertainty estimation, clear communication of limitations, and participatory
design with local stakeholders to avoid overreliance on automated outputs.

\section{Conclusion}
We documented a LiteFNO reproduction on the Gray-Scott reaction--diffusion
dataset with full configuration transparency and quantitative results. The
current run delivers strong validation and test metrics with a compact model,
and provides a foundation for expanded sweeps and baseline comparisons in
future low-resource deployment studies.

\bibliographystyle{plain}
\bibliography{references}

\end{document}
